Staged Abstract Interpreters

### 2019-08-16 paper/10_intro.tex
Moar proofreading

Abstract interpretation as a lattice-based approach to sound static
analyses was proposed by \citet{DBLP:conf/popl/CousotC77}. Equipped
with Galois connections \cite{CousotCousot79-1}, the analyzer can
soundly approximate the program runtime behavior by computing fixed
points on an abstract domain. Despite the tremendous theoretical
development of abstract interpretation over the years, constructing
artifacts and analyzers that perform sound abstract interpretation for
modern and expressive languages was considered complicated for a long
time.

@@

Based on the idea of abstracting abstract machines, more
pragmatically, several implementation strategies utilizing purely
functional programming techniques to build abstract interpreters have
emerged. Such techniques include monads, monad transformers
\cite{DBLP:journals/pacmpl/DaraisLNH17,
  Sergey:2013:MAI:2491956.2491979}, arrows
\cite{Keidel:2018:CSP:3243631.3236767}, extensible effects
\cite{Kiselyov:2015:FMM:2804302.2804319}, etc. The pure approaches provide certain
benefits, including that the abstract interpretation artifacts can be
built compositionally and modularly, e.g., by using monad
transformers. Therefore, the soundness of analysis can be proven with
less effort, whether by mechanized proofs
\cite{Darais:2016:CGC:2951913.2951934} or paper-based proofs
\cite{Keidel:2018:CSP:3243631.3236767}.  Also, referential
transparency and equational reasoning allow programmers to
confidently reason about the correctness of their implementations.

However, besides the intrinsic complexity of static analysis, there
are additional abstraction penalties from these high-level
implementation approaches. First, similar to concrete interpreters,
an abstract interpreter analyzes the program by traversing the
abstract syntax tree (AST), which incurs interpretive overhead, such
as pattern matching on the ASTs and recursive calls on the
subexpressions.
Such overhead can be negligible if the abstract interpreter only runs
on the program a single time, but can accumulate significantly if
run repeatedly (for example, on libraries).  Second, it is common that
abstract interpreters written in pure languages extensively use effect
systems to implement the semantics of abstract interpretation. For
example, the collecting interpreter that returns all possible runtime
values is a form of nondeterminism, where nondeterminism monads can
help.  Although such pure approaches have their own merits, compared
with imperative stateful implementations, they are significantly
slower.

>>>>

Abstract interpretation as a lattice-based approach to sound static analyses was
proposed by \citet{DBLP:conf/popl/CousotC77}. Based on the notion of Galois
connections \cite{CousotCousot79-1}, the analyzer can soundly approximate the
concrete program runtime behavior by computing fixed points on an abstract
domain. Despite the tremendous theoretical development of abstract
interpretation over the years, constructing artifacts and analyzers that perform
sound abstract interpretation for modern and expressive languages was considered
complicated for a long time.

@@

Based on the idea of abstracting abstract machines, more
pragmatically, several implementation strategies utilizing purely
functional programming techniques to build abstract interpreters have
emerged. Such techniques include monads, monad transformers
\cite{DBLP:journals/pacmpl/DaraisLNH17,
  Sergey:2013:MAI:2491956.2491979}, arrows
\cite{Keidel:2018:CSP:3243631.3236767}, extensible effects
\cite{Githubsemantic}, etc. The pure approaches provide certain
benefits, including that the abstract interpretation artifacts can be
built compositionally and modularly, e.g., by using monad
transformers. Therefore, the soundness of analysis can be proven with
less effort, whether by mechanized proofs
\cite{Darais:2016:CGC:2951913.2951934} or paper-based proofs
\cite{Keidel:2018:CSP:3243631.3236767}.  Also, referential
transparency and equational reasoning allow programmers to more
confidently reason about the correctness of their implementations.

However, besides the intrinsic complexity of static analysis, there
are additional abstraction penalties with these high-level
implementation approaches. First, similar to concrete interpreters,
an abstract interpreter analyzes the program by traversing the
abstract syntax tree (AST), which incurs an interpretive overhead, such
as the pattern matching on the ASTs and recursive calls on the
subexpressions.
Such overhead can be negligible if the abstract interpreter only runs
on the program a single time, but can accumulate significantly if
run repeatedly (for example, on libraries).  Second, it is common that
abstract interpreters written in pure languages extensively use effect
systems to implement the semantics of abstract interpretation. For
example, the abstract interpreter that returns all possible runtime
values is a form of nondeterminism, where nondeterminism monads can
help.  Although such pure approaches have their own merits, compared
with imperative stateful implementations, they are significantly
slower.

### 2019-08-16 paper/abstract.tex
Moar proofreading

It is well known that a staged interpreter is a compiler: specializing an
interpreter to a given program produces an equivalent executable that runs faster.
This connection is known as the first Futamura projection.
It is even more widely known that an abstract interpreter is a program analyzer:
tweaking an interpreter to run on an abstract domain produces a sound static
analysis. What happens when we combine these two ideas, and apply specialization to
an \emph{abstract} interpreter?

In this paper, we present a unifying framework that naturally extends the first
Futamura projection from concrete interpreters to abstract interpreters. Our
approach derives a sound staged abstract interpreter based on a
semantic-agnostic interpreter with type-level binding-time abstraction and
monadic abstraction. By using different instantiations of these abstractions,
the generic interpreter can flexibly behave in four modes: unstaged concrete
interpreter, staged concrete interpreter, unstaged abstract interpreter, or
staged abstract interpreter.
As an example of \emph{abstraction without regret}, the overhead of these
abstraction layers is eliminated in the generated code after staging.
We show that staging abstract interpreters is practical and useful to
optimize static analysis while requiring less engineering efforts without 
compromising soundness. We conduct three case studies, including a comparison
with \citeauthor{Boucher:1996:ACN:647473.727587}'s abstract compilation,
applications on various control-flow analyses, and a demonstration that
the approach can be used to compile an analysis to the program in a modular
fashion.  We also empirically evaluate the running time improved by staging.
The experiment shows an order of magnitude speedup with staging for
control-flow analyses.

>>>>

It is well known that a staged interpreter is a compiler: specializing an
interpreter to a given program produces an equivalent executable that runs faster.
This connection is known as the first Futamura projection.
It is even more widely known that an abstract interpreter is a program analyzer:
tweaking an interpreter to run on an abstract domain produces a sound static
analysis. What happens when we combine these two ideas, and apply staging to
an \emph{abstract} interpreter?

In this paper, we present a unifying framework that naturally extends the first
Futamura projection from concrete interpreters to abstract interpreters. Our
approach derives a sound staged abstract interpreter based on a
semantic-agnostic interpreter with type-level binding-time abstraction and
monadic abstraction. By using different instantiations of these abstractions,
the generic interpreter can flexibly behave in four modes: unstaged concrete
interpreter, staged concrete interpreter, unstaged abstract interpreter, or
staged abstract interpreter. As an example of \emph{abstraction without regret},
the overhead of these abstraction layers is eliminated in the generated code
after staging. We show that staging abstract interpreters is practical and
useful to optimize static analysis while requiring less engineering effort,
without compromising soundness. We conduct three case studies, including a
comparison with \citeauthor{Boucher:1996:ACN:647473.727587}'s abstract
compilation, applications to various control-flow analyses, and a demonstration
that can be used for modular analysis. We also empirically evaluate the effect
of staging on the execution time. The experiment shows an order of magnitude
speedup with staging for control-flow analyses.
